---
title: "Programming basics"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
checker <- function(label, user_code, check_code, envir_result, evaluate_result, ...) {
  list(message = check_code, correct = TRUE, location = "append")
}
tutorial_options(exercise.timelimit = 60, exercise.checker = checker)
```



## Welcome 

### Welcome to R

R is easiest to use when you understand the R language. This tutorial will teach you the essential background knowledge necessary for every piece of R code. You'll learn about:

* **functions** and their **arguments**
* **objects**
* R's basic **data types** 
* R's basic data structures including **vectors** and **lists**
* R's **package system**

## Functions 

### Functions

To begin learning about __functions__, please watch  ![](https://vimeo.com/220490105) from the RStudio Education team.

### Run a function

Can you use the `sqrt()` function in the chunk below to compute the square root of 961?

```{r functions-1, exercise = TRUE}

```

```{r functions-1-solution}
sqrt(961)
```

### Code

Use the code chunk below to examine the code that `sqrt()` runs.

```{r functions-2, exercise = TRUE}

```

```{r functions-2-solution}
sqrt
```

```{r functions-2-check}
"Good job! sqrt immediately triggers a low-level algorithm optimized for performance, so there is little code to see."
```

### lm

Compare the code in `sqrt()` to the code in another R function, `lm()`. Examine `lm()`'s code body in the chunk below.

```{r functions-3, exercise = TRUE}

```

```{r functions-3-solution}
lm
```

### help pages

Wow! `lm()` runs a lot of code. What does it do? Open the help page for `lm()` in the chunk below and find out.

```{r functions-4, exercise = TRUE}

```

```{r functions-4-solution}
?lm
```

```{r functions-4-check}
"Good job! R's `lm()` function is used for fitting basic linear models. It's no wonder that it runs so much code. "
```

### Code comments

What do you think the chunk below will return? Run it and see. The result should be nothing. R will not run anything on a line after a `#` symbol. This is useful because it lets you write human-readable comments in your code: place them after a `#`. Now delete the `#` and re-run the chunk. You should see a result.

```{r functions-5, exercise = TRUE}
# sqrt(961)
```

```{r functions-5-solution}
sqrt(961)
```


## Arguments

### Arguments

To begin learning about function __arguments__, please watch  ![](https://vimeo.com/220490157) from the RStudio Education team.

### args()

`rnorm()` is a function that generates random variables from a normal distribution. Find the arguments of `rnorm()` using the `args()` function.

```{r arguments-1, exercise = TRUE}

```

```{r arguments-1-solution}
args(rnorm)
```

```{r arguments-1-check}
"Good job! `n` specifies the number of random normal variables to generate. `mean` and `sd` describe the distribution with which to generate the random values. We will learn more about a distribution's mean and standard deviation later in the course." 
```

### optional arguments

```{r arguments-2, echo = FALSE}
question("Which arguments of R norm are optional?",
         answer("n"),
         answer("mean", correct = TRUE),
         answer("sd", correct = TRUE),
         allow_retry = TRUE,
         correct = "The only required argument is `n` because it lacks a default value.")
```

### rnorm() 

Use `rnrom()` to generate 100 random normal values with a mean of 100 and a standard deviation of 15.

```{r arguments-3, exercise = TRUE}

```

```{r arguments-3-solution}
rnorm(100, mean = 100, sd = 15)
```

### rnorm(), again

Can you spot the error in the code below? Fix the code and then re-run it.

```{r arguments-4, exercise = TRUE}
rnorm(100, mu = 100, sd = 15)
```

```{r arguments-4-solution}
rnorm(100, mean = 100, sd = 15)
```

## Objects

### Objects

To begin learning about __objects__, please watch  ![](https://vimeo.com/220493412) from the RStudio Education team.

### Object names

You can choose almost any name you like for an object, as long as the name does not begin with a number or a special character like `+`, `-`, `*`, `/`, `^`, `!`, `@`, or `&`. 

```{r objects-1, echo = FALSE}
question("Which of these would be valid object names?",
  answer("today", correct = TRUE),
  answer("1st"),
  answer("+1"),
  answer("vars", correct = TRUE),
  answer("\\^_^"),
  answer("foo", correct = TRUE),
  allow_retry = TRUE,
  correct = "It's important to give your objects descriptive names that will help you remember what they are."
)
```

### Using objects

In the code chunk below, save the results of `rnorm(100, mean = 100, sd = 15)` to an object named `data`. Use the assignment operator `<-` to save the results in the object names `data`. Then, on a new line, call the `plot()` function on `data` to plot the random values against a sequence from 1 to 100.

```{r objects-2, exercise = TRUE}

```

```{r objects-2-solution}
data <- rnorm(100, mean = 100, sd = 15)
plot(data)
```

### What if?

What do you think would happen if you assigned `data` to a new object named `copy`, like this? Run the code and then inspect both `data` and `copy`.

```{r objects-3, exercise = TRUE}
data <- rnorm(100, mean = 100, sd = 15)
copy <- data
```

```{r objects-3-solution}
data <- rnorm(100, mean = 100, sd = 15)
copy <- data
data
copy
```

```{r objects-3-check}
"Good job! R saves a copy of the contents in data to copy."
```

### Datasets

Objects provide an easy way to store datasets in R. In fact, R comes with many toy datasets pre-loaded. Examine the contents of `iris` to see a classic toy dataset.    
_Hint:_ How could you learn more about the `iris` object?

```{r objects-4, exercise = TRUE}

```

```{r objects-4-solution}
iris
```

```{r objects-4-check}
"Good job! You can learn more about iris by examining its help page with `?iris`."
```

### rm()

What if you accidentally overwrite an object? If that object came with R or one of its packages, you can restore the original version of the object by removing your version with `rm()`. Run `rm()` on `iris` below to restore the iris dataset.

```{r objects-5, exercise = TRUE}
iris <- 1
iris
```

```{r objects-5-solution}
iris <- 1
iris
rm(iris)
iris
```

```{r objects-5-check}
"Great work! However, please remember that using the `rm()` function won't be helpful if you accidentally overwrite one of your own objects."
```

## Vectors

### Vectors

To begin learning about __vectors__, please watch  ![](https://vimeo.com/220490316) from the RStudio Education team.

### Create a vector

In the chunk below, use the function `c()` to create a vector that contains the integers from one to ten.

```{r vectors-1, exercise = TRUE}

```

```{r vectors-1-solution}
c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
```

### : (colon operator)

If your vector contains a sequence of contiguous integers, you can create it with the `:` shortcut. Run `1:10` in the chunk below. What do you get? What do you suppose `1:20` would return?

```{r vectors-2, exercise = TRUE}

```

```{r vectors-2-solution}
1:10
```

### [ ]

You can extract any element from a vector by placing a pair of square brackets after the vector's name. Simply include the number of the element you want to extract inside the brackets. For example, `vec[3]` will return the third element of the vector named `vec`.

Use the chunk below to extract the fourth element of `vec`.

```{r vectors-3, exercise = TRUE}
vec <- c(1, 2, 4, 8, 16)
```

```{r vectors-3-solution}
vec <- c(1, 2, 4, 8, 16)
vec[4]
```

### More [ ]

You can also use `[]` to extract multiple elements of a vector. Place the vector `c(1,2,5)` between the brackets below. What does R return?

```{r vectors-4, exercise = TRUE}
vec <- c(1, 2, 4, 8, 16)
vec[]
```

```{r vectors-4-solution}
vec <- c(1, 2, 4, 8, 16)
vec[c(1,2,5)]
```

### Names

If the elements of your vector have names, you can extract them by name. Place a name or vector of names in the square brackets behind the vector object to do so. Surround each name with quotation marks, e.g., `vec2[c("alpha", "beta")]`.

Extract the element named gamma from the vector below.

```{r vectors-5, exercise = TRUE}
vec2 <- c(alpha = 1, beta = 2, gamma = 3)
```

```{r vectors-5-solution}
vec2 <- c(alpha = 1, beta = 2, gamma = 3)
vec2["gamma"]
```

### Vectorized operations

Predict what the code below will return. Then, look at the result.

```{r vectors-6, exercise = TRUE}
c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10) + c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
```

```{r vectors-6-check}
"Good job! Like many R functions, R's math operators are vectorized: they're designed to work with vectors by repeating the operation for each pair of elements."
```

### Vector recycling

Predict what the code below will return. Then, look at the result.

```{r vectors-7, exercise = TRUE}
1 + c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
```

```{r vectors-7-check}
"Good job! Whenever you try to work with vectors of varying lengths (recall that `1` is a vector of length one), R will repeat the shorter vector as needed to compute the result."
```

## Types

### Types

To begin learning about __data types__, please watch  ![](https://vimeo.com/220490241) from the RStudio Education team.

### Atomic types

```{r types1, echo = FALSE}
question("Which of these is not an atomic data type",
         answer("numeric/double"),
         answer("integer"),
         answer("character"),
         answer("logical"),
         answer("complex"),
         answer("raw"),
         answer("simple", correct = TRUE),
         allow_retry = TRUE,
         random_answer_order = TRUE)
```

### What type?

```{r types2, echo = FALSE}
question('What type of data is `"1L"`',
         answer("numeric/double"),
         answer("integer", message = "Close. `1L` would be an integer, but look closer."),
         answer("character", correct = TRUE),
         answer("logical"),
         allow_retry = TRUE)
```

### Integers

Create a vector of integers from one to five. Can you imagine why you might want to use integers instead of numbers/doubles?

```{r types3, exercise = TRUE}

```

```{r types3-solution}
c(1L, 2L, 3L, 4L, 5L)
```

```{r types3-check}
"Good job! An integer vector uses less memory compared to double vector, having the same values. Compare  the result of object.size(c(1L, 2L, 3L, 4L, 5L)) with that of object.size(c(1, 2, 3, 4, 5))."
```


### Floating point arithmetic

Computers have a limited amount of memory to store decimal numbers, which can sometimes require infinite precision. As a result, some decimal values can only be represented as very precise approximations. Occasionally, you may notice the effects of this imprecision, as shown below.

Calculate the square root of two, then square that result (multiply the square root of two by itself), and subtract two from the final answer. What outcome do you expect? What result do you obtain?

```{r types4, exercise = TRUE}

```

```{r types4-solution}
sqrt(2) * sqrt(2) - 2
sqrt(2)^2 - 2
```

### Vectors

```{r types5, echo = FALSE}
question('How many types of data can you put into a single vector?',
         answer("1", correct = TRUE),
         answer("6"),
         answer("As many as you like"),
         allow_retry = TRUE)
```

### Character or object?

One of the most common mistakes in R is to call an object when you mean to call a character string and vice versa.

```{r types6, echo = FALSE}
question('Which of these are object names? What is the difference between object names and character strings?',
         answer("foo", correct = TRUE),
         answer('"num"'),
         answer("mu", correct = TRUE),
         answer('"sigma"'),
         answer('"data"'),
         answer("a", correct = TRUE),
         allow_retry = TRUE,
         correct = "Character strings are surrounded by quotation marks, object names are not.")
```

## Lists 

### Lists

To begin learning about __lists__, please watch  ![](https://vimeo.com/220490360) from the RStudio Education team.

### Lists vs. vectors

```{r lists1, echo = FALSE}
question('Which data structure(s) could you use to store these pieces of data in the same object? `1001`, `TRUE`, `"stories"`.',
         answer("a vector"),
         answer("a list", correct = TRUE),
         answer("neither"),
         allow_retry = TRUE)
```

### Make a list

Make a list that contains the elements `1001`, `TRUE`, and `"stories"`. Give each element a name.

```{r lists2, exercise = TRUE}

```

```{r lists2-solution}
list(number = 1001, logical = TRUE, string = "stories")
```

### Extract an element

Extract the number 1001 from the list below.

```{r lists3, exercise = TRUE}
things <- list(number = 1001, logical = TRUE, string = "stories")
```

```{r lists3-solution}
things <- list(number = 1001, logical = TRUE, string = "stories")
things$number
```

### Data Frames

You can make a data frame with the `data.frame()` function, which works similarly to `c()` and `list()`. Assemble the vectors below into a data frame with the column names `numbers`, `logicals`, and `strings`.

```{r lists4, exercise = TRUE}
nums <- c(1, 2, 3, 4)
logs <- c(TRUE, TRUE, FALSE, TRUE)
strs <- c("apple", "banana", "carrot", "duck")
```

```{r lists4-solution}
nums <- c(1, 2, 3, 4)
logs <- c(TRUE, TRUE, FALSE, TRUE)
strs <- c("apple", "banana", "carrot", "duck")
data.frame(numbers = nums, logicals = logs, 
           strings = strs)
```

```{r lists4-check}
"Good Job. To create a data frame, it is essential that each column vector is of equal length."
```

### Extract a column

To extract the strings column from the `df` data frame, keep in mind that a data frame is a special type of list with named elements. Do it!

```{r lists5, exercise = TRUE}
nums <- c(1, 2, 3, 4)
logs <- c(TRUE, TRUE, FALSE, TRUE)
strs <- c("apple", "banana", "carrot", "duck")
df <- data.frame(numbers = nums, logicals = logs, strings = strs)
```

```{r lists5-solution}
df$strings
```

## Packages

### Packages

To begin learning about __packages__, please watch  ![](https://vimeo.com/220490447) from the RStudio Education team.

### A common error


```{r packages1, echo = FALSE}
question('What does this common error message suggest? `object _____ does not exist.`',
         answer("You misspelled your object name"),
         answer("You've forgot to load the package that ____ comes in"),
         answer("Either", correct = TRUE),
         allow_retry = TRUE)
```

### Load a package

In the code chunk below, load the `tidyverse` package. Whenever you load a package R will also load all the packages that the first package depends on.

```{r packages2, exercise = TRUE}

```

```{r packages2-solution}
library(tidyverse)
```

```{r packages2-check}
"Good job! R will keep the packages loaded until you close your R session. When you re-open R, you'll need to reload you packages."
```

### Quotes

Did you know, `library()` is a special function in R? You can pass `library()` a package name in quotes, like `library("tidyverse")`, or not in quotes, like `library(tidyverse)`---both will work! That's often not the case with R functions.

In general, you should always use quotes unless you are writing the _name_ of something that is already loaded into R's memory, like a function, vector, or data frame.

### Install packages

But what if the package that you want to load is not installed on your computer? How would you install the dplyr package on your own computer?

```{r packages3, exercise = TRUE}

```

```{r packages3-solution}
install.packages("dplyr")
```

```{r packages3-check}
"Good job! You only need to install a package once, unless you wish to update your local copying by reinstalling the package. Notice that `install.packages()` _always_ requires quotes around the package name."
```

## Tidyverse

### The tidyverse package

The ["tidyverse"](http://tidyverse.org) is a collection of packages designed for data science, sharing a common philosophy that enables them to work well together. 


### Installing the tidyverse


It's easy to install all packages from the **tidyverse**. You simply need to run

```{r eval = FALSE}
install.packages("tidyverse")
```

This command tells R to install all tidyverse packages simultaneously. When you run `install.packages("tidyverse")`, R will install the following packages in one simple step:

* ggplot2
* dplyr
* tidyr
* readr
* purrr
* tibble
* hms
* stringr
* lubridate
* forcats
* DBI
* haven
* jsonlite
* readxl
* rvest
* xml2
* modelr
* broom
* ...

### loading the tidyverse

When you load the **tidyverse** with `library("tidyverse")`, it instructs R to load _the most commonly used_ tidyverse packages. These are:

* ggplot2
* dplyr
* tidyr
* readr
* purrr
* tibble
* forcats
* lubridate
* stringr

You can load the less commonly used tidyverse packages in the usual way by running `library(<PACKAGE NAME>)` for each of them.

Let's give this a try. We will use the ggplot2 and dplyr packages later in this tutorial. Let's use the tidyverse package to load them in the chunk below:

```{r packages4, exercise = TRUE}

```

```{r packages4-solution}
library("tidyverse")
```

### Quiz

```{r packages5, echo = FALSE}
question('Which package is not loaded by `library("tidyverse")`',
  answer("ggplot2"),
  answer("dplyr"),
  answer("tibble"),
  answer("babynames", correct = TRUE, message = "Now that you are familiar with the data set, and have loaded the necessary packages, let's explore the data."),
  allow_retry = TRUE
)
```

### Congratulations!

Congratulations! You now have a formal understanding of the basics of R. Whenever R behaves unexpectedly, you can use your knowledge of how R works to determine what went wrong.
